---
title: "DEG ANALYSIS"
author: "VAIBHAVI"
date: "2025-04-03"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r}
install.packages("knitr")
install.packages("rmarkdown")
```

#Prepare DEG analysis
```{r}

setwd("C:/Users/Student/Desktop/Count_Practice/Count_Practice")

```


#Load count data
```{r}
install.packages("gtools")
library("gtools")
filelist <- mixedsort(list.files(pattern = "*.txt"))
data_list <- lapply(filelist, function(fn) read.delim(fn, header = FALSE, sep = "\t")[, 2])
countdata <- do.call(cbind, data_list)
rownames(countdata) <- read.delim(filelist[1], header = FALSE, sep = "\t")[, 1]
```


#Polish count data
```{r}
colnames(countdata) <-sub(".count.txt", "",filelist)
mode(countdata) <-"numeric"
countdata <- countdata[-c((nrow(countdata)-4):nrow(countdata)),]
head(countdata)
tail(countdata)
View(countdata)
```

#Generate colData
```{r}
sample.names <- colnames(countdata)
coldata <- data.frame(  
name = sample.names,        
condition = rep(c("control", "treated"), each = 2, times = 2),   
sample = rep(c("A", "B"), each = 4) )


coldata$condition <- factor(coldata$condition)
coldata$sample <- factor(coldata$sample)

coldata

write.csv(coldata,file="coldata.csv",row.names = FALSE)

```


#Prepare DESeq2 data
```{r}
install.packages("DESeq2")
library(DESeq2)
dds <- DESeqDataSetFromMatrix(
countData = countdata,
colData = coldata,
design = ~ condition + sample)
```


#Plot dispersion of genes
```{r}
dds <- DESeq(dds)
plotDispEsts(dds,main="Dispersion plot")
```

#Activity
```{r}
dds <- DESeq(dds)
plotDispEsts(dds,main="Dispersion plot",
 xlab = "Normalized counts",
 ylab = "Dispersion level",
 genecol = "orange",
 fitcol = "red",
  finalcol = "blue",
 cex = 0.30,)
 
```




#DEG analysis practice
#Normalization using DESeq2
```{r}
dds.SF <-estimateSizeFactors(dds)
print(sizeFactors(dds.SF))
```


```{r}
Normalizedcount <- counts(dds.SF, normalized = T)
head(Normalizedcount)
```

```{r}

write.csv(Normalizedcount, file = "Normalized_counts.csv", row.names = TRUE)

```

```{r}
dds.SF <-estimateSizeFactors(dds)
print(sizeFactors(dds.SF))

Normalizedcount <- counts(dds.SF, normalized = T)
head(Normalizedcount)

write.csv(Normalizedcount, file="Normalized_counts.csv", row.names = TRUE)
```




#Plot counts
```{r}
group.col <- c("brown","red")[coldata$condition]
png(filename = "plot_counts.png", width = 12, height = 6, units = "in",  res = 300 )
par(mfrow=c(1,2), mar = c(5, 10, 5, 5))

boxplot(log2(counts(dds.SF, normalized=FALSE)+1),
xlab="log2(counts+1)", main="Raw counts" , las = 1, 
horizontal = TRUE, pch = ",", col = group.col)

boxplot(log2(counts(dds.SF, normalized=TRUE)+1), 
xlab="log2(normalized counts+1)", main="Normalized counts", las = 1, 
horizontal = TRUE, pch = ",", col = group.col)

dev.off()

```



#Sample distance examination
```{r}
BiocManager::install("gplots")
library("gplots")
library("ggplot2")

transformed_data <- vst(dds, blind = TRUE)

sampleDists <- dist(t(assay(transformed_data)))

sampleDistMatrix <- as.matrix(sampleDists)
```




#Heatmap
```{r}

colours <- colorRampPalette(c("red", "white"))(100)


heatmap(  
sampleDistMatrix,  
Rowv = NA,   
Colv = NA,  
col = colours,  
margins = c(12, 12),  
main = "Sample distance"  
)

```




#Principal Component Analysis (PCA)
```{r}
pcaData <- plotPCA(transformed_data, intgroup = c("sample","condition"),returnData = TRUE)

percentVar <-round(100 * attr(pcaData,"pecentVar"))

PCA <- ggplot(pcaData, aes(PC1, PC2, color = condition,shape = sample))+
   geom_point(size = 3, alpha = 0.8) +
  xlab(paste0("PC1 (", percentVar[1], "%) ")) +
  ylab(paste0("PC2 (", percentVar[2], "%) ")) +
scale_color_manual(values = c("skyblue", "orange")) 

PCA

ggsave("../PCA_plot.png", plot = PCA, width = 5, height = 4, dpi = 300)
```




#Activity
```{r}
PCA <- ggplot(pcaData, aes(PC1, PC2, color = condition,shape = sample))+
geom_point(size = 3, alpha = 0.8) +
xlab(paste0("PC1 (",percentVar[1],"%) ")) +
ylab(paste0("PC2 (",percentVar[2],"%) ")) +
theme_bw() +
ggtitle("PCAPLOT") +
scale_color_manual(values = c("blue", "red")) +
  
theme(plot.title =element_text(size = 12, face ="bold", hjust = 0.5),
legend.title=element_text(size=12),
legend.text=element_text(size=12)) +
  
theme(panel.grid.major = element_blank()) +
theme(panel.grid.minor = element_blank())
  
scale_color_manual(values = c("skyblue", "orange")) 

PCA


ggsave("../PCA_plot.png", plot = PCA, width = 5, height = 4, dpi = 300)
```




#Filtering
```{r}

ddsFiltered <- dds

filterGenes <- rowSums(counts(ddsFiltered, normalized = TRUE) >= 10) >= 2

ddsFiltered <- ddsFiltered[filterGenes, ]

plotDispEsts(ddsFiltered, main= "Dispersion after filtering") 

```




#Run DESeq function
```{r}

design(ddsFiltered) <- ~ sample + condition + sample:condition

ddsFiltered <- DESeq(ddsFiltered)

resultsNames(ddsFiltered)

resSampleA <- results(ddsFiltered, name = "condition_treated_vs_control")
summary(resSampleA)

resSampleB <- results(ddsFiltered, name = "sampleB.conditiontreated")
summary(resSampleB)

```



#Extract significant DEGs
```{r}
sigDEGs_A <- subset(resSampleA, padj<0.05 & abs(log2FoldChange)> 1)
sigDEGs_A
write.csv(as.data.frame(sigDEGs_A),file = "SampleA_DEGs.csv")


sigDEGs_B <- subset(resSampleB, padj < 0.05 & abs(log2FoldChange) > 1)
sigDEGs_B
write.csv(as.data.frame(sigDEGs_B), file = "SampleB_DEGs.csv")

```




#Visualization and GO analysis
#volcano plot
```{r}
resSampleA <- resSampleA[!is.na(resSampleA$padj), ]


resSampleA$Significance <- ifelse(resSampleA$padj < 0.05 & abs(resSampleA$log2FoldChange) > 1, "Significant", "Not Significant")


library(ggplot2)
volcanoA <- ggplot(resSampleA, aes(x = log2FoldChange, y = -log10(padj), color = Significance)) +  
geom_point(alpha = 0.8, size = 1.5) +  
scale_color_manual(values = c("Significant" = "green", "Not Significant" = "blue")) +  
labs(title = "Sample A",       
x = "Log2 Fold Change",      
y = "-Log10 Adjusted P-value")

volcanoA

```



#Activity
```{r}
resSampleB <- resSampleB[!is.na(resSampleB$padj), ]


resSampleB$Significance <- ifelse(resSampleB$padj < 0.05 & abs(resSampleB$log2FoldChange) > 1, "Significant", "Not Significant")


library("ggplot2")
volcanoB <- ggplot(resSampleB, aes(x = log2FoldChange, y = -log10(padj), color = Significance)) +  
geom_point(alpha = 0.8, size = 1.5) +  
scale_color_manual(values = c("Significant" = "yellow", "Not Significant" = "black")) +
  labs(title = "Sample B",       
x = "Log2 Fold Change",      
y = "-Log10 Adjusted P-value") +
  theme_bw() +
  theme(
    legend.position = "top",
    axis.title = element_text(size = 17),   
    axis.text = element_text(size = 12),   
    panel.grid.major = element_blank(),     
    panel.grid.minor = element_blank()      
)


volcanoB

```


#volcano plot sample A
```{r}
resSampleA <- resSampleA[!is.na(resSampleA$padj), ]


resSampleA$Significance <- ifelse(resSampleA$padj < 0.05 & abs(resSampleA$log2FoldChange) > 1, "Significant", "Not Significant")


library(ggplot2)
volcanoA <- ggplot(resSampleA, aes(x = log2FoldChange, y = -log10(padj), color = Significance)) +  
geom_point(alpha = 0.8, size = 1.5) +  
scale_color_manual(values = c("Significant" = "red", "Not Significant" = "yellow")) +
  labs(title = "Sample A",       
x = "Log2 Fold Change",      
y = "-Log10 Adjusted P-value") +
  theme_bw() +
  theme(
    legend.position = "top",
    axis.title = element_text(size = 17),   
    axis.text = element_text(size = 12),   
    panel.grid.major = element_blank(),     
    panel.grid.minor = element_blank()      
)


volcanoA

```




#Plot specific genes
#Prepare normalized count data with coldata.
```{r}
BiocManager::install("Rmisc")
library(Rmisc)

NormalizedCount <- read.csv("Normalized_counts.csv", row.names = 1)
NC_t <- cbind(name = rownames(t(NormalizedCount)), t(NormalizedCount))


coldata <- read.csv("coldata.csv", header = TRUE, row.names = 1)
coldata$name <- rownames(coldata)


NC_col <- merge(coldata, NC_t, by = "name")


NC_col[, 4:ncol(NC_col)] <- lapply(NC_col[, 4:ncol(NC_col)], function(x) as.numeric(as.character(x)))

```


#Summarize data for a gene
```{r}

Glyma_13G122766.sum <- summarySE(NC_col, measurevar = "Glyma_13G122766", groupvars = c("sample", "condition"))


Glyma_13G122766.sum

```


#Plotting
```{r}

Glyma_13G122766 <- ggplot(Glyma_13G122766.sum, aes(x = sample, y = Glyma_13G122766, fill = condition)) +
geom_bar(position = position_dodge(), stat = "identity", color = "grey20") +
  labs(y = "Transcript levels\n(normalized counts)", 
       x = "Sample", 
       title = "Glyma_13G122766") +
  geom_errorbar(aes(ymin = Glyma_13G122766, ymax = Glyma_13G122766 + sd),
 linewidth = 0.2, color = "red", width = 0.1, position = position_dodge(0.5)) +
  scale_fill_manual(values = c("blue", "black"),
                    breaks = c("control", "treated"))

Glyma_13G122766  


ggsave(filename = "Glyma_13G122766.png",
       plot = Glyma_13G122766, 
       width = 2, height = 3, units = "in", dpi = 300)


```


#Activity
##Summarize data for a gene
```{r}

Glyma_01G000359.sum <- summarySE(NC_col, measurevar = "Glyma_01G000359", groupvars = c("sample", "condition"))


Glyma_01G000359.sum

```


#Plotting
```{r}

Glyma_01G000359 <- ggplot(Glyma_01G000359.sum, aes(x = sample, y = Glyma_01G000359, fill = condition)) +
geom_bar(position = position_dodge(), stat = "identity", color = "blue") +
  labs(y = "Transcript levels\n(normalized counts)", 
       x = "Sample", 
       title = "Glyma_01G000359") +
  geom_errorbar(aes(ymin = Glyma_01G000359, ymax = Glyma_01G000359 + sd),
 linewidth = 0.3, color = "orange", width = 0.2, position = position_dodge(0.9)) +
  scale_fill_manual(values = c("red", "black"),
                    breaks = c("control", "treated"))

Glyma_01G000359


ggsave(filename = "Glyma_01G000359.png",
       plot = Glyma_01G000359, 
       width = 4, height = 4, units = "in", dpi = 300)

```






